---
title: "bayesian_reproduction"
format: html
editor: visual
date: 2026-03-18
author: Anna Rosa Klaver, Are Markku Alanen Tjihkkom, Jemima Agnew, Liradie Allen, Naomi Takizawa
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))
```

# Article

Hostetter, A. B., & Bahl, S. (2023). Comparing the cognitive load of gesture and action production: A dual-task study. *Language and Cognition*, *15*(3), 601–621. <https://doi.org/10.1017/langcog.2023.23>

# Buddy Challenge Task 4

*Run a series of Bayesian models equivalent to the critical model in the original article that originally did not converge:*

(a) *run the closest equivalent of their analysis (i.e. same structure but using flat priors)*
(b) *run a proper model with proper, weakly informative priors and justify your prior choices; interpret the results relative to the research question by interpreting the posteriors relative to zero; write up the results and write a report on the comparison between your reanalysis and their original analysis. Write at least 500 words in which you describe your analysis including your prior choices and compare the results between your analysis and the original analysis both quantitatively and qualitatively (do the analyses arrive at the same conclusions?). add the report to your GitHub Repo from Part 3.*

# Data Preparation

Repeating our data-tidying from the analytical reproduction, using a new copy of the data CSV.

## Load the data

```{r}
#| label: load-gesture

# Load a copy of the original data into a dataframe
gesture <- read_csv('bayesian_repro_data.csv')

# Have a look
head(gesture)
```

## Tidy the data

```{r}
#| label: tidy-gesture

# Exclude the data in which participant didn't follow the instructions
gesture_tidy <- gesture |>
  filter(Include == 'y')

# Remove columns we will not use
gesture_tidy <- gesture_tidy |>
  select(- c(Script,
             `Index Words`,
            `Start Time`,
              Transcript,
         `End Time`,
         Comment,
         Include,
         TimeTalking)) |>
  # Make the column names more stylistically consistent
  # And make them better reflect terminology used in the article
  rename(Item = Trial, # We are guessing they use 'trial' to record specific combinations of grid and dot pattern, which they refer to when modeling as 'item'
         Deictic = index.speech,
         Time_speaking = Seconds,
         Word_count = wordcount,
         Filled_pauses = Disfluenices,
         Letters_recalled = Letters,
         Locations_recalled = Locations,
         Total_recalled = Both) |>
  # Make condition labels clearer
  mutate(Condition = case_when(Condition == 'M' ~ 'Make',
                               Condition == 'G' ~ 'Gesture',
                               Condition == 'NG' ~ 'Hands still'),
         # Convert letters and locations recalled to numeric, so we can predict them
         # with a linear model later
         Letters_recalled = as.numeric(Letters_recalled),
         Locations_recalled = as.numeric(Locations_recalled),
         # make condition a factor, for difflsmeans later
         Condition = as.factor(Condition))
```

# a) Model with flat priors

Focusing on the final part of the article's study 1 analysis, since it actually addresses their main research question.

## Get priors

```{r}
#| label: model formulas

# set model formulas for final two models

letters_formula <- bf(Letters_recalled ~                          
                       Condition +                          
                       Filled_pauses +                          
                       (1 + Condition | Participant) +                      
                       (1 + Condition | Item))  


locations_formula <- bf(Locations_recalled ~                            
                          Condition +                            
                          Filled_pauses +                            
                          Deictic +                            
                          (1 + Condition | Participant) +                       
                          (1 + Condition | Item))  
# check priors 
get_prior(letters_formula, gesture_tidy)  

get_prior(locations_formula, gesture_tidy)

  # priors are functionally identical for both models, so we will be using one set of priors for letters and locations models 
```

## Model

```{r}
#| label: models1 

# specify uniform priors 
priors_flat <- c(# System would prefer we specify argument `lb` & `ub` to make explicit lower and upper bounds. Also for class `b`.
                    prior(uniform(0,5), class  = 'Intercept'),  
                    # `b` priors are flat by default apparently, but still
                    prior(uniform(-5, 5), class = 'b'),
                    # `sd` priors describe random effects
                    prior(uniform(-5, 5), class = 'sd'),
                    # This also seems to be the default, but we state it explicitly
                    prior(lkj(1), class = 'cor'))  

# run letters model 
letters_mdl_flat <- brm(letters_formula,                    
                   family = gaussian(),                    
                   prior = letters_priors,                    
                   data = gesture_tidy,
                   file = 'models/letters_mdl_flat.RDS',
                   # MCMC specs:                    
                   chains = 4, cores = 4,                    
                   iter = 2000, warmup = 1000,                    
                   thin = 1, seed = 16)  

# model output 
letters_mdl_flat

# run locations model 
locations_mdl_flat <- brm(locations_formula,                    
                     family = gaussian(),                    
                     prior = locations_priors,                    
                     data = gesture_tidy,    
                     file = 'models/locations_mdl_flat.RDS',
                     # MCMC specs:                    
                     chains = 4, cores = 4,                    
                     iter = 2000, warmup = 1000,                    
                     thin = 1, seed = 16)  

# model output 
locations_mdl_flat
```

# b) Model with weakly informative priors

## Set priors and run model

```{r}
#| label: models3

# Set weakly informative priors for intercept and predictors.
priors_informative <- 
  c(prior(normal(2.5, 2.5), class = 'Intercept'),
    prior(normal(0, 2.5), class = 'b'),
    prior(normal(0, 2.5), class = 'sd'),
    prior(lkj(1), class = 'cor'),
    prior(normal(0, 2.5), class = "sigma"))
  # Text copied from Jo:
  # We do not manually specify priors for the fixed effects (class = "b");
  # brms defaults are the closest practical analogue to flat/diffuse priors here.
  # Random-effect standard deviations need proper priors in a Bayesian MLM.
  # A broad Student-t prior is weakly informative and computationally safer than a bounded uniform prior.
  
  # Residual standard deviation also needs a proper positive prior.
 # Again, we use a broad weakly informative Student-t.
  

# Run bayesian model with weakly informative priors
letters_model_informative <-
  brm(letters_formula,
       family = gaussian(),
       prior = priors_informative,
       data = gesture_tidy,
       # Save model in a nearby file
       file = "models/letters_mdl_informative.RDS",
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)

letters_model_informative

locations_model_informative <- 
  brm(locations_formula,
       family = gaussian(),
       prior = priors_informative,
       data = gesture_tidy,
       # Save model in a nearby file
       file = "models/locations_mdl_informative.RDS",
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)

locations_model_informative
```

## Interpret the results

### Checks and posterior plotting

```{r}
#| label: checks

pp_check(letters_model_informative, ndraws = 100)

pp_check(locations_model_informative, ndraws = 100)

mcmc_plot(letters_model_informative, type = 'trace')
mcmc_plot(locations_model_informative, type = 'trace')

# Shit goes kinda flimpy over there.

mcmc_dens_overlay(letters_model_informative)

# No. Not legible. For our human eyes anyway.

p_direction(letters_model_informative)
p_direction(locations_model_informative)

# Plot the posteriors

letters_model_informative |> 
  spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses) |>
  gather_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses) |> 
  ggplot(aes(x = .value)) +
  stat_halfeye(fill = '#756bb1') +
  geom_vline(xintercept = 0, lty = 'dashed') +
  geom_density(fill = "#bcbddc", alpha = 0.5) +
  facet_wrap(~ .variable, scales = "free") +
  labs(
    x = "\nPosterior slope",
    y = "Density\n",
    title = "Distribution of posteriors for model predicting letter recall\n") +
  theme_minimal()

locations_model_informative |> 
  spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic) |>
  gather_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic) |> 
  ggplot(aes(x = .value)) +
  stat_halfeye(fill = '#756bb1') +
  geom_vline(xintercept = 0, lty = 'dashed') +
  geom_density(fill = "#bcbddc", alpha = 0.5) +
  facet_wrap(~ .variable, scales = "free") +
  labs(
    x = "\nPosterior slope",
    y = "Density\n",
    title = "Distribution of posteriors for model predicting location recall\n") +
  theme_minimal()
```

### Bayes Factor

```{r}
#| label: bayes_factors 

# Extract posteriors

letters_posteriors <- letters_model_informative |> 
  spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses) |>
  select(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses)

locations_posteriors <- locations_model_informative |> 
  spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic) |>
  select(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic)

# Create prior models

letters_model_prior <-
  brm(letters_formula,
       family = gaussian(),
       prior = priors_informative,
       sample_prior = 'only',
       data = gesture_tidy,
       # Save model in a nearby file
       file = "models/letters_mdl_prior.RDS",
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)

locations_model_prior <- 
  brm(locations_formula,
       family = gaussian(),
       prior = priors_informative,
       sample_prior = 'only',
       data = gesture_tidy,
       # Save model in a nearby file
       file = "models/locations_mdl_prior.RDS",
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)

# Extract priors

letters_priors <- letters_model_prior |> 
  spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses) |>
  select(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses)

locations_priors <- locations_model_prior |> spread_draws(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic) |>
  select(b_ConditionHandsstill, b_ConditionMake, b_Filled_pauses, b_Deictic)

# Bayes Factor for letters model
bayesfactor_parameters(letters_posteriors, letters_priors, direction = "two-sided", null = 0)

# Bayes Factor for locations model
bayesfactor_parameters(locations_posteriors, locations_priors, direction = "two-sided", null = 0)

```

For the model predicting correct recall of letters (0-5 recalled correctly), the 'condition' predictor has a base level of 'gesture', and alternate levels of 'hands still' and 'make'. The Bayes factor for hands still as opposed to gesture is 0.581, indicating anecdotal evidence favoring the null hypothesis that there is no effect of hands still condition. The Bayes factor for make is 0.062, indicating strong evidence favoring the null. The 'filled pauses' predictor is numeric (ranging from 0-13), using the number of filled pauses recorded as a predictor of letter recall. Its Bayes factor is 0.015, indicating very strong evidence favoring the null that there is no effect of filled pauses.

For the model predicting correct recall of locations (0-5 recalled correctly), condition and filled pauses predictor are also used. The Bayes factor for hands still as opposed to gesture is 0.046, indicating strong evidence favoring the null hypothesis that there is no effect of hands still condition. The Bayes factor for make is 0.065, indicating strong evidence favoring the null. The Bayes factor for filled pauses is 0.027, indicating very strong evidence favoring the null that there is no effect of filled pauses. This model also uses number of deictic references as a predictor (ranging from 0-1), which has a Bayes factor of 1.45, indicating anecdotal evidence favoring the alternative hypothesis that there is an effect of deictic references on locations recalled.

### Write-up

*Task: write a report on the comparison between your reanalysis and their original analysis. Write at least 500 words in which you describe your analysis including your prior choices and compare the results between your analysis and the original analysis both quantitatively and qualitatively (do the analyses arrive at the same conclusions?). Add the report to your GitHub Repo from Part 3.*

# Buddy Challenge Task 5:

# Hypothesis testing using ROPEs

*Define the smallest effect size of interest for the original research question and do a literature research to justify your choice.*

-   *rerun your hypothesis test combining the concepts of Bayes Factor (BF) and Regions of Practical Equivalence (ROPE) (according to Roettger & Franke's method, 2025)*

-   *run a sensitivity analyses for different ROPES and priors; write up the results accordingly*

-   *save the analysis and report in your Github repository*

```{r}
#| label: ROPEs

# set a ROPE
rope <- c(-0.5, 0.5)

# calculate Bayes Factor using the ROPE
  # for the letters model
bayesfactor_rope(posterior = letters_model_informative,
                 prior = letters_model_prior,
                 null = rope,
                 parameters = c('b_ConditionHandsstill', # < 0.01
                                'b_ConditionMake', # < 0.01
                                'b_Filled_pauses')) # < 0.01

  # for the locations model
bayesfactor_rope(posterior = locations_model_informative,
                 prior = locations_model_prior,
                 null = rope,
                 parameters = c('b_ConditionHandsstill', # < 0.01
                                'b_ConditionMake', # < 0.01
                                'b_Filled_pauses', # < 0.01
                                'b_Deictic')) # ~ 0.01
```

## Sensitivity analysis

### Exploring varying priors

```{r}
#| label: models with varying priors

# DO NOT RUN THIS CHUNK, MODELS HAVE ALREADY BEEN SAVED
# Repeat our priors for all parameters except the predictors (class 'b')
priors <- c(prior(normal(2.5, 2.5), class = 'Intercept'),
            prior(normal(0, 2.5), class = 'sd'),
            prior(lkj(1), class = 'cor'),
            prior(normal(0, 2.5), class = "sigma"))

# five different prior widths that are reasonable
priors_1.5 <- c(priors, prior(normal(0, 1.5), class = 'b'))
priors_2 <- c(priors, prior(normal(0, 2), class = 'b'))
# our original prior width
priors_2.5 <- c(priors, prior(normal(0, 2.5), class = 'b'))
priors_3 <- c(priors, prior(normal(0, 3), class = 'b'))
priors_3.5 <- c(priors, prior(normal(0, 3.5), class = 'b'))

# Define a list of these prior specifications
prior_list <- list(priors_1.5,
                   priors_2,
                   priors_2.5,
                   priors_3,
                   priors_3.5)

#initialise lists to store models
letters_model_list <- list()
locations_model_list <- list()

# Loop over the priors for letters models
for (i in seq_along(prior_list)) {
  letters_model_list[[i]] <-
     brm(letters_formula,
       family = gaussian(),
       prior = prior_list[[i]],
       data = gesture_tidy,
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)
}

# name models
names(letters_model_list) <- paste0("full_model_with_prior_", c(1.5,2,2.5,3,3.5))
# save models
saveRDS(letters_model_list, "models/letters_model_loop.RDS")


# Loop over the priors for locations models
for (i in seq_along(prior_list)) {
  locations_model_list[[i]] <-
     brm(locations_formula,
       family = gaussian(),
       prior = prior_list[[i]],
       data = gesture_tidy,
       chains = 4, cores = 4, 
       iter = 2000, warmup = 1000,
       thin = 1, seed = 16)
}

# name models
names(locations_model_list) <- paste0("full_model_with_prior_", c(1.5,2,2.5,3,3.5))
# save models
saveRDS(locations_model_list, "models/locations_model_loop.RDS")
```

```{r}
#| label: loading in the model lists

# load models into new objects
letters_model_list <- readRDS("models/letters_model_loop.RDS")
locations_model_list <- readRDS("models/locations_model_loop.RDS")
```

### Exploring varying ROPEs

```{r}
#| label: exploring different ROPEs

# specify different ROPE intervals of interest
ropes <- list(c(-0.1, 0.1),
              c(-0.3, 0.3),
              c(-0.5, 0.5),
              c(-0.7, 0.7),
              c(-1, 1))

# loop through different ropes: LETTERS
letters_results <- map_dfr(names(letters_model_list), function(model_name) {
  model <- letters_model_list[[model_name]]
  
  map_dfr(ropes, function(interval) {
    lower <- interval[1]
    upper <- interval[2]
    
    bayesfactor_parameters(
      model,
      null = c(lower, upper),
      parameters = c('b_ConditionHandsstill', 
                                'b_ConditionMake', 
                                'b_Filled_pauses')
    ) %>%
      as_tibble() %>%
      mutate(
        model = model_name,
        interval_range = paste0("[", lower, ", ", upper, "]")
      ) %>%
      select(model, everything())
  })
})

# wrangle per parameter (I've ignored the results for filled pauses because that's not relevant for the hypotheses)
letters_results_handsstill <- letters_results |> 
  filter(Parameter == "b_ConditionHandsstill") |> 
  separate(model, sep = "_", into = c(NA, NA, NA, NA, "prior_sd")) |> 
  mutate(BF = exp(log_BF)) |> 
  select(prior_sd, BF, interval_range)

letters_results_make <- letters_results |> 
  filter(Parameter == "b_ConditionMake") |> 
  separate(model, sep = "_", into = c(NA, NA, NA, NA, "prior_sd")) |> 
  mutate(BF = exp(log_BF)) |> 
  select(prior_sd, BF, interval_range)

# loop through different ropes: LOCATIONS
locations_results <- map_dfr(names(locations_model_list), function(model_name) {
  model <- locations_model_list[[model_name]]
  
  map_dfr(ropes, function(interval) {
    lower <- interval[1]
    upper <- interval[2]
    
    bayesfactor_parameters(
      model,
      null = c(lower, upper),
      parameters = c('b_ConditionHandsstill', 
                                'b_ConditionMake', 
                                'b_Filled_pauses',
                                'b_Deictic')
    ) %>%
      as_tibble() %>%
      mutate(
        model = model_name,
        interval_range = paste0("[", lower, ", ", upper, "]")
      ) %>%
      select(model, everything())
  })
})

# wrangle per parameter (I've ignored the results for fixed pauses because that's not relevant for the hypotheses)
locations_results_handsstill <- locations_results |> 
  filter(Parameter == "b_ConditionHandsstill") |> 
  separate(model, sep = "_", into = c(NA, NA, NA, NA, "prior_sd")) |> 
  mutate(BF = exp(log_BF)) |> 
  select(prior_sd, BF, interval_range)

locations_results_make <- locations_results |> 
  filter(Parameter == "b_ConditionMake") |> 
  separate(model, sep = "_", into = c(NA, NA, NA, NA, "prior_sd")) |> 
  mutate(BF = exp(log_BF)) |> 
  select(prior_sd, BF, interval_range)
```

```{r}
#| label: sensitivity analysis plots

# visualise the BFs for the letter models in a raster
letters_handsstill_plot <- ggplot(letters_results_handsstill,
       aes(x = as.factor(prior_sd),
           y = as.factor(interval_range),
           fill = BF)) +
  geom_tile(colour = "grey") +
  geom_text(aes(label = round(BF,3))) +
  scale_fill_gradient2(transform = "log",
                       limits = c(0.1,10),
                       midpoint = 1, 
                       low = "#5C7457", 
                       high = "#FA8100",
                       breaks = c(0.00000,0.03,0.1,0.33,0,3,10),
                       labels = c("0","0.03","0.1","0.33","1","3","10"),
                       oob = scales::squish
                       ) +
  labs(title = "(A) Bayes factor in favour of null", 
       subtitle = "Effect of hands still condition on letter recall",
       fill = "BF",
       y = "ROPE in semi tones centered on 0\n ",
       x = "\nprior standard deviation") +
  theme_minimal()

letters_handsstill_plot

letters_make_plot <- ggplot(letters_results_make,
       aes(x = as.factor(prior_sd),
           y = as.factor(interval_range),
           fill = BF)) +
  geom_tile(colour = "grey") +
  geom_text(aes(label = round(BF,3))) +
  scale_fill_gradient2(transform = "log",
                       limits = c(0.1,10),
                       midpoint = 1, 
                       low = "#5C7457", 
                       high = "#FA8100",
                       breaks = c(0.00000,0.03,0.1,0.33,0,3,10),
                       labels = c("0","0.03","0.1","0.33","1","3","10"),
                       oob = scales::squish
                       ) +
  labs(title = "(A) Bayes factor in favour of null", 
       subtitle = "Effect of make condition on letter recall",
       fill = "BF",
       y = "ROPE in semi tones centered on 0\n ",
       x = "\nprior standard deviation") +
  theme_minimal()

letters_make_plot

# visualise the BFs for the location models in a raster
locations_handsstill_plot <- ggplot(locations_results_handsstill,
       aes(x = as.factor(prior_sd),
           y = as.factor(interval_range),
           fill = BF)) +
  geom_tile(colour = "grey") +
  geom_text(aes(label = round(BF,3))) +
  scale_fill_gradient2(transform = "log",
                       limits = c(0.1,10),
                       midpoint = 1, 
                       low = "#5C7457", 
                       high = "#FA8100",
                       breaks = c(0.00000,0.03,0.1,0.33,0,3,10),
                       labels = c("0","0.03","0.1","0.33","1","3","10"),
                       oob = scales::squish
                       ) +
  labs(title = "(A) Bayes factor in favour of null", 
       subtitle = "Effect of hands still condition on location recall",
       fill = "BF",
       y = "ROPE in semi tones centered on 0\n ",
       x = "\nprior standard deviation") +
  theme_minimal()

locations_handsstill_plot

locations_make_plot <- ggplot(letters_results_make,
       aes(x = as.factor(prior_sd),
           y = as.factor(interval_range),
           fill = BF)) +
  geom_tile(colour = "grey") +
  geom_text(aes(label = round(BF,3))) +
  scale_fill_gradient2(transform = "log",
                       limits = c(0.1,10),
                       midpoint = 1, 
                       low = "#5C7457", 
                       high = "#FA8100",
                       breaks = c(0.00000,0.03,0.1,0.33,0,3,10),
                       labels = c("0","0.03","0.1","0.33","1","3","10"),
                       oob = scales::squish
                       ) +
  labs(title = "(A) Bayes factor in favour of null", 
       subtitle = "Effect of make condition on location recall",
       fill = "BF",
       y = "ROPE in semi tones centered on 0\n ",
       x = "\nprior standard deviation") +
  theme_minimal()

locations_make_plot
```
