---
title: "bayesian_reproduction"
format: html
editor: visual
date: 2026-03-18
author: Anna Rosa Klaver, Are Markku Alanen Tjihkkom, Jemima Agnew, Liradie Allen, Naomi Takizawa
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))
```

# Article

Hostetter, A. B., & Bahl, S. (2023). Comparing the cognitive load of gesture and action production: A dual-task study. *Language and Cognition*, *15*(3), 601–621. <https://doi.org/10.1017/langcog.2023.23>

# Task

Run a series of Bayesian models equivalent to the critical model in the original article that originally did not converge:

(a) run the closest equivalent of their analysis (i.e. same structure but using flat priors)
(b) run a proper model with proper, weakly informative priors and justify your prior choices; interpret the results relative to the research question by interpreting the posteriors relative to zero; write up the results and write a report on the comparison between your reanalysis and their original analysis. Write at least 500 words in which you describe your analysis including your prior choices and compare the results between your analysis and the original analysis both quantitatively and qualitatively (do the analyses arrive at the same conclusions?). add the report to your GitHub Repo from Part 3.

# Data Preparation

Repeating our data-tidying from the analytical reproduction, using a new copy of the data CSV.

## Load the data

```{r}
#| label: load-gesture

# Load a copy of the original data into a dataframe
gesture <- read_csv('bayesian_repro_data.csv')

# Have a look
head(gesture)
```

## Tidy the data

```{r}
#| label: tidy-gesture

# Exclude the data in which participant didn't follow the instructions
gesture_tidy <- gesture |>
  filter(Include == 'y')

# Calculate proportion of data excluded
prop_excluded <- 1 - (nrow(gesture_tidy)/nrow(gesture))
prop_excluded
# This gets 10.8%, which is slightly more then the 10.6% the authors reported.
# We have discarded 127 trials, which is 2 more than the 125 they report in total.

# Remove columns we will not use
gesture_tidy <- gesture_tidy |>
  select(- c(Script,
             `Index Words`,
            `Start Time`,
              Transcript,
         `End Time`,
         Comment,
         Include,
         TimeTalking)) |>
  # Make the column names more stylistically consistent
  # And make them better reflect terminology used in the article
  rename(Item = Trial, # We are guessing they use 'trial' to record specific combinations of grid and dot pattern, which they refer to when modeling as 'item'
         Deictic = index.speech,
         Time_speaking = Seconds,
         Word_count = wordcount,
         Filled_pauses = Disfluenices,
         Letters_recalled = Letters,
         Locations_recalled = Locations,
         Total_recalled = Both) |>
  # Make condition labels clearer
  mutate(Condition = case_when(Condition == 'M' ~ 'Make',
                               Condition == 'G' ~ 'Gesture',
                               Condition == 'NG' ~ 'Hands still'),
         # Convert letters and locations recalled to numeric, so we can predict them
         # with a linear model later
         Letters_recalled = as.numeric(Letters_recalled),
         Locations_recalled = as.numeric(Locations_recalled),
         # make condition a factor, for difflsmeans later
         Condition = as.factor(Condition))
```

# a) run the closest equivalent of their analysis (i.e. same structure but using flat priors)

Focusing on the final part of the article's study 1 analysis, since it actually addresses their main research question.

## Models predicting recall by condition

### Letters Model: Uniform priors

```{r}
#| label: models1 
#| 
# full random effects structure for the model selected in the final part of the og analysis 

letters_formula <- bf(Letters_recalled ~                          
                       Condition +                          
                       Filled_pauses +                          
                       (1 + Condition | Participant) +                      
                       (1 + Condition | Item))  
# check priors 
get_prior(letters_formula, gesture_tidy)  

# specify uniform priors 
priors_flat <- c(# System would prefer we specify argument `lb` & `ub` to make explicit lower and upper bounds. Also for class `b`.
                    prior(uniform(0,5), class  = 'Intercept'),  
                    # `b` priors are flat by default apparently, but still
                    prior(uniform(-5, 5), class = 'b'),
                    # `sd` priors describe random effects
                    prior(uniform(-5, 5), class = 'sd'),
                    # This also seems to be the default, but we state it explicitly
                    prior(lkj(1), class = 'cor'))  

# run model 
letters_mdl_flat <- brm(letters_formula,                    
                   family = gaussian(),                    
                   prior = letters_priors,                    
                   data = gesture_tidy,
                   file = 'models/letters_mdl_flat.RDS',
                   # MCMC specs:                    
                   chains = 4, cores = 4,                    
                   iter = 2000, warmup = 1000,                    
                   thin = 1, seed = 16)  

# model output 
letters_mdl_flat
```

### Locations Model: Uniform priors

```{r}
#| label: models2 

# full random effects structure for the model selected in the final part of the og analysis 
locations_formula <- bf(Locations_recalled ~                            
                          Condition +                            
                          Filled_pauses +                            
                          Deictic +                            
                          (1 + Condition | Participant) +                       
                          (1 + Condition | Item))  

# run model 
locations_mdl_flat <- brm(locations_formula,                    
                     family = gaussian(),                    
                     prior = locations_priors,                    
                     data = gesture_tidy,    
                     file = 'models/locations_mdl_flat.RDS',
                     # MCMC specs:                    
                     chains = 4, cores = 4,                    
                     iter = 2000, warmup = 1000,                    
                     thin = 1, seed = 16)  

# model output 
locations_mdl_flat
```

# b) run a proper model with proper, weakly informative priors and justify your prior choices; interpret the results relative to the research question by interpreting the posteriors relative to zero; write up the results and write a report on the comparison between your reanalysis and their original analysis. Write at least 500 words in which you describe your analysis including your prior choices and compare the results between your analysis and the original analysis both quantitatively and qualitatively (do the analyses arrive at the same conclusions?). add the report to your GitHub Repo from Part 3.

```{r}
#| label: models3

# Set weakly informative priors for intercept and predictors.
priors_letters_informative <- c(
  # Text copied from Jo:
  # We do not manually specify priors for the fixed effects (class = "b");
  # brms defaults are the closest practical analogue to flat/diffuse priors here.
  # Random-effect standard deviations need proper priors in a Bayesian MLM.
  # A broad Student-t prior is weakly informative and computationally safer than a bounded uniform prior.
  prior(student_t(df = 3, location = 0, scale = 2.5,
        class = 'sd')),
  # Residual standard deviation also needs a proper positive prior.
 # Again, we use a broad weakly informative Student-t.
  prior(student_t(df = 3, location = 0, scale = 2.5), class = "sigma"))
  

# Run bayesian model with weakly informative priors
letters_model_informative <-
  brm(letter_formula,
       family = gaussian(),
       prior = flat_priors_letter,
       data = gesture_tidy,
       # Save model in a nearby file
       file = "models/letter_mdl_informative.RDS",
       chains = 4,
       cores = 4, 
       iter = 2000,
       warmup = 1000,
       thin = 1,
       seed = 42)
```
