---
title: "study_1_reproduction"
format: html
editor: visual
date: 2026-02-13
author: Anna Rosa Klaver, Are Markku Alanen Tjihkkom, Jemima Agnew, Liradie Allen, Naomi Takizawa
execute:
  error: false
  warning: false
  message: false
  cache: false
#bibliography: 
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, lmerTest)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))
```

# Article

Hostetter, A. B., & Bahl, S. (2023). Comparing the cognitive load of gesture and action production: A dual-task study. *Language and Cognition*, *15*(3), 601–621. <https://doi.org/10.1017/langcog.2023.23>

# Task

Assess whether you can analytically reproduce their analyses including descriptive numbers, graphs, and inferential results according to what they report in the paper (not following their scripts but their description in the paper).

# Data Preparation

## Load the data

```{r}
#| label: load-gesture

# Load a copy of the original data into a dataframe
gesture <- read_csv('reproduction_AGI_study_1_data.csv')

# Have a look
head(gesture)
```

## Tidy the data

```{r}
#| label: tidy-gesture

# Exclude the data in which participant didn't follow the instructions
gesture_tidy <- gesture |>
  filter(Include == 'y')

# Calculate proportion of data excluded
prop_excluded <- 1 - (nrow(gesture_tidy)/nrow(gesture))
prop_excluded
# This gets 10.8%, which is slightly more then the 10.6% the authors reported.
# We have discarded 127 trials, which is 2 more than the 125 they report in total.

# Remove columns we will not use
gesture_tidy <- gesture_tidy |>
  select(- c(Script,
             `Index Words`,
            `Start Time`,
              Transcript,
         `End Time`,
         Comment,
         Include,
         TimeTalking)) |>
  # Make the column names less awful
  rename(Item = Trial, # We are guessing they use 'trial' to record specific combinations of grid and dot pattern, which they refer to when modeling as 'item'
         Deictic = index.speech,
         Word_count = wordcount,
         Filled_pauses = Disfluenices,
         Letters_recalled = Letters,
         Locations_recalled = Locations,
         Total_recalled = Both) |>
  # Make condition labels clearer
  mutate(Condition = case_when(Condition == 'M' ~ 'Make',
                               Condition == 'G' ~ 'Gesture',
                               Condition == 'NG' ~ 'Hands still'))
```

## Check data aggregates

```{r}
#| label: gesture-aggregates

# Calculate total number of participants remaining
length(unique(gesture_tidy$Participant))
# 49 matches the number reported.

# Determine the trial counts for each condition x participant combination
trial_counts_conditional <- table(Condition = gesture_tidy$Condition,
                                  Participant = gesture_tidy$Participant) |>
  as.data.frame()
head(trial_counts_conditional)

# Find the aggregate means and ranges per condition
trial_counts_conditional |>
  group_by(Condition) |>
  summarise(mean_trials_per_participant = mean(Freq),
            min_trials_per_participant = min(Freq),
            max_trials_per_participant = max(Freq))

# Why is there a min 0 from make? Let's see which participant has this number.
trial_counts_conditional |>
  group_by(Condition, Participant) |>
  summarise(mean_trials_per_participant = mean(Freq),
            min_trials_per_participant = min(Freq),
            max_trials_per_participant = max(Freq))
# The one case with zero trials per participant is Make condition, participant 2.
# This likely is why our overall mean for Make is lower than the 6.50 they report.
# But the mean trials per participant were spot-on otherwise, so I don't think they dropped participant 2. This likely connects to the different exclusion numbers we got.
```

# Models to predict speech measure by condition

# Run models

```{r}
#| label: mixed models

word_count_mdl <- lmer(Word_count ~ Condition +
                         (1 + Condition | Participant) +
                         (1 + Condition | Item),
                       data = gesture_tidy)
# This converged.

time_speaking_mdl <- lmer(Seconds ~ Condition +
                         (1 + Condition | Participant) +
                         (1 + Condition | Item),
                       data = gesture_tidy)
# Singular fit.
summary(time_speaking_mdl)
# We don't see a clear reason why this is singular.
# Remove random slopes for participant, as reported.
time_speaking_mdl_reduced <- lmer(Seconds ~ Condition +
                         (1 | Participant) +
                         (1 | Item),
                       data = gesture_tidy)
# No longer singular

filled_pauses_mdl <- lmer(Filled_pauses ~ Condition +
                         (1 + Condition | Participant) +
                         (1 + Condition | Item),
                       data = gesture_tidy)
# Singular
filled_pauses_mdl_reduced <- lmer(Filled_pauses ~ Condition +
                         (1 | Participant) +
                         (1 | Item),
                       data = gesture_tidy)
# Not singular

deictic_mdl <- glmer(Deictic ~ Condition +
                         (1 + Condition | Participant) +
                         (1 + Condition | Item),
                       data = gesture_tidy,
                       family = 'binomial')
# Actually did not converge. Dropping random slopes
deictic_mdl <- glmer(Deictic ~ Condition +
                         (1 | Participant) +
                         (1 | Item),
                       data = gesture_tidy,
                       family = 'binomial')
# This worked.
```

The maximal binomial model to predict deictic reference got the following error when it failed to converge:

`Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :   unable to evaluate scaled gradient Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :   Model failed to converge: degenerate  Hessian with 1 negative eigenvalues`

## Test significance of pairwise comparisons between levels of `condition`

```{r}
#| label: model_comparison

# Using the ls_means function was not specified in the article, and we were confused. We cheated and checked the writers' script.
table_yay <- ls_means(word_count_mdl)
difflsmeans(word_count_mdl)
# These functions do not show us a table with any values.
```

So far, we cannot test pairwise differences based on these models.
